---
title: "Graphs_4500"
output: github_document
date: "2025-04-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Below is my graphs for our debt project. The graph below is looking at a comparison between people of color and white people for median debt collections. Median debt collections is the average amount that people are behind on payments.

```{r}
library(ggplot2)
library(dplyr)
library(readxl)
library(reshape2)
library(tidyr)

#import the data
Debt_2024 <- read_excel("dia_lbls_all_overall_usa_2023_1Jul2024.xlsx")



#rename the data
names(Debt_2024) <- c("country", "share_with_debt_all", "share_with_debt_color", "share_with_debt_white", "median_debt_all", "median_debt_color", "median_debt_white", "medical_debt_all", "medical_debt_color", "medical_debt_white", "student_all", "student_color", "student_white", "auto_all", "auto_color", "auto_white", "creditcard_del_all", "creditcard_del_color", "creditcard_del_white", "median_creditcard_all", "median_creditcard_color", "median_creditcard_white", "share_color", "avg_house_income_all", "avg_house_income_color", "avg_house_income_white")

#create a subset for the thing you want to look at
subset_debt_2024 <- select(Debt_2024, median_debt_all, median_debt_color, median_debt_white)
names(subset_debt_2024) <- c("All_People","People_of_Color", "White_People")


#create another subset in the long format
subset_debt_2024_long <- tidyr::pivot_longer(subset_debt_2024, cols = everything(), names_to = "group", values_to = "median_debt")



#create the plot
ggplot(subset_debt_2024_long, aes(x = group, y = median_debt, fill = group)) +
  geom_bar(stat = "identity") +
  labs(title = "Comparison of Median Debt in Collection in the US", subtitle = "The average amount of money people are behind on payments", y = "Median Debt in Collections") +
  theme_minimal()
```

```{r}
#create another graph from same data set

subset_income_2024 <- select(Debt_2024, avg_house_income_all, avg_house_income_color, avg_house_income_white)
names(subset_income_2024) <- c("All_People","People_of_Color", "White_People")


#create another subset in the long format
subset_income_2024_long <- tidyr::pivot_longer(subset_income_2024, cols = everything(), names_to = "group", values_to = "avg_household_income")


subset_income_2024_long$avg_household_income <- as.numeric(as.character(subset_income_2024_long$avg_household_income))

#create the plot
ggplot(subset_income_2024_long, aes(x = group, y = avg_household_income, fill = group)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Household Income by Group",
       x = "Group",
       y = "Average Income") +
  theme_minimal()
```


#student debt by state

```{r}
student_loan_debt_2023 <- read_excel("dia_lbls_all_student_state_2023_1Jul2024.xlsx")

names(student_loan_debt_2023) <- c("state_FIPS", "state_name", "state_abbreviation", "share_with_studentloan_debt_all", "share_with_studentloan_debt_color", "share_with_studentloan_debt_white", "median_studentloan_debt_all", "median_studentloan_debt_color", "median_studentloan_debt_white", "share_of_holders_with_debt_all", "share_of_holders_with_debt_color", "share_of_holders_with_debt_white", "median_loan_debt_in_default_all", "median_loan_debt_in_default_color", "median_loan_debt_in_default_white", "median_monthly_studentloan_payment_all", "median_monthly_studentloan_payment_color", "median_monthly_studentloan_payment_white", "share_without_bachelors_all", "share_without_bachelors_color", "share_without_bachelors_white", "share_of_people_of_color", "average_household_income_all", "average_household_income_color", "average_household_income_white")

subset_student_loan_debt <- select(student_loan_debt_2023, state_FIPS, state_name, state_abbreviation, median_studentloan_debt_all, median_studentloan_debt_color, median_studentloan_debt_white)

subset_student_loan_debt$median_studentloan_debt_all <- as.numeric(as.character(subset_student_loan_debt$median_studentloan_debt_all))

#sort the data in descending order and find top 5 states with the highest median student loan debt
sorted_student_loan_data <- subset_student_loan_debt |> arrange(desc(median_studentloan_debt_all))

top_five_states_with_student_loan_debt <- head(sorted_student_loan_data, 5)

#change format to long

top_five_student_loans_long <- top_five_states_with_student_loan_debt |>
  pivot_longer(cols = c(median_studentloan_debt_white, median_studentloan_debt_color), names_to = "Group", values_to = "Median_Loan")


top_five_student_loans_long$Median_Loan <- as.numeric(as.character(top_five_student_loans_long$Median_Loan))

#create a bar chart comparing these top 5 states

ggplot(top_five_student_loans_long, aes(x = state_name, y = Median_Loan, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Median Student Loans by State",
 x = "State",
 y = "Median Student Loan",
 fill = "Group") +
  scale_y_continuous(limits = c(0, NA)) +
 theme_minimal()

```

Code for coloring the bars based off of people group (DONT USE)
```{r}
melted_data <- melt(top_five_states_with_student_loan_debt, id.vars = "state_name", 
                    measure.vars = c("median_studentloan_debt_color", "median_studentloan_debt_white"),
                    variable.name = "Group", value.name = "Median_Student_Loan_Debt")

melted_data$Median_Student_Loan_Debt <- as.numeric(as.character(melted_data$Median_Student_Loan_Debt))


ggplot(melted_data, aes(x = state_name, y = Median_Student_Loan_Debt, fill = Group)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = Median_Student_Loan_Debt), position = position_stack(vjust = 0.5), size = 3) +
  labs(title = "Median Student Loan Debt for Top Five States",
       x = "State",
       y = "Median Student Loan Debt") +
  scale_fill_manual(values = c("median_studentloan_debt_color" = "skyblue", "median_studentloan_debt_white" = "orange"),
                    labels = c("Colored People", "White People")) +
  theme_minimal()
```

Now lets do the states with the most difference between colored people and white people
```{r}
#first create a column for the difference between median for colored people and white people
subset_student_loan_debt$median_studentloan_debt_color <- as.numeric(as.character(subset_student_loan_debt$median_studentloan_debt_color))

subset_student_loan_debt$median_studentloan_debt_white <- as.numeric(as.character(subset_student_loan_debt$median_studentloan_debt_white))

subset_student_loan_debt$Difference <- subset_student_loan_debt$median_studentloan_debt_color - subset_student_loan_debt$median_studentloan_debt_white

subset_student_loan_debt$Difference <- as.numeric(as.character(subset_student_loan_debt$Difference))


subset_student_loan_debt$Difference <- abs(subset_student_loan_debt$Difference)


sorted_data_for_biggest_diff <- subset_student_loan_debt |> arrange(desc(Difference))

top_five_states_difference <- head(sorted_data_for_biggest_diff, 5)

melted_data_difference <- melt(top_five_states_difference, id.vars = "state_name", 
                    measure.vars = c("median_studentloan_debt_color", "median_studentloan_debt_white"),
                    variable.name = "Group", value.name = "Median_Student_Loan_Debt_Diff")



#real graph
ggplot(melted_data_difference, aes(x = state_name, y = Median_Student_Loan_Debt_Diff, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Median Student Loans by State", subtitle = "Top 5 states with the greatest difference between people groups",
 x = "State",
 y = "Median Student Loan",
 fill = "Group") +
  scale_y_continuous(limits = c(0, NA)) +
 theme_minimal()
```

This next section is looking at some statistics by race over time. This graph is for percent of people holding student loans

```{r}
library(readr)
long_term_student_debt <- read_csv("interactive_bulletin_charts_racecl4_have.csv")

long_term_student_debt$year <- as.numeric(long_term_student_debt$year)
long_term_student_debt$Education_Installment_Loans <- as.numeric(long_term_student_debt$Education_Installment_Loans)

#create the graph for percent holding student loans


ggplot(long_term_student_debt, aes(x = year, y = Education_Installment_Loans, color = Category, shape = Category)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Percent Holding Student Loan Debt by Race or Ethnicity",
       x = "Year",
       y = "% Holding Installment Loans") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(1989, 2022, by = 3)) +
  scale_y_continuous(breaks = seq(0, max(long_term_student_debt$Education_Installment_Loans), by = 5))



```

Looking at more trends over time. This one is median $ in thousands of dollars of student loan debt

```{r}
data_over_time_student_median <- read_csv("interactive_bulletin_charts_racecl4_median.csv")

data_over_time_student_median$year <- as.numeric(data_over_time_student_median$year)
data_over_time_student_median$Education_Installment_Loans <- as.numeric(data_over_time_student_median$Education_Installment_Loans)

#create graph
ggplot(data_over_time_student_median, aes(x = year, y = Education_Installment_Loans, color = Category, shape = Category)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Median Education Installment Loans by Race or Ethnicity",
       x = "Year",
       y = "Median Student Loans (thousands of $)") +
  theme_minimal() +
  theme(legend.title = element_blank()) +
  scale_x_continuous(breaks = seq(1989, 2022, by = 3)) +
  scale_y_continuous(breaks = seq(0, max(data_over_time_student_median$Education_Installment_Loans), by = 5))
```

